\myfont Implementation and evaluation of register allocation for translating LLVM-{}- to x86-64

### 2024-01-09 article.tex
Explain between example and tables
Add linearscan as well as brief explanations

Consider a simple program like Figure \ref{fig:countc} that only prints integers incrementally up until \texttt{argc}, in LLVM-{}-  that would consist of one loop with one phi node as seen in Figure \ref{fig:count1ll}. The number of iterations is nearly halfed when reverseing the list of instructions prior to dataflow as can be seen on \ref{tab:countll-rev-steps} when compared to \ref{tab:countll-steps}.

@@

@@

\begin{figure}[H]
     \centering
   \begin{minipage}[b]{0.50\linewidth}
     \centering
     \includesvg[height=10.9em]{countdown.briggs2}
     \caption{\texttt{build -{}- -t dot -a briggs -n 2}}\label{fig:countdown.briggs.dot2}
   \end{minipage}
   \begin{minipage}[b]{0.32\linewidth}
     \centering
     \includesvg[height=9em]{countdown.briggs}
     \caption{\texttt{-a briggs -n 3}}\label{fig:countdown.briggs.dot}
   \end{minipage}
\end{figure}

@@

The primary text is \textit{Linear Scan Register Allocation} \cite{linear}, the application of which also relies on prior liveness analysis. Given liveranges, assignments are easily derived from only one pass through IR.
Similar to graph coloring, given \(K\) available GPRs, it must assign  as many variables as possible to these, and spill the remaining.

>>>>

Consider a simple program like Figure \ref{fig:countc} that only prints integers incrementally up until \texttt{argc}, in LLVM-{}-  that would consist of one loop with one phi node as seen in Figure \ref{fig:count1ll}.

@@

\noindent If one were to note the \(\mathit{in}[s]\) and \(\mathit{out}[s]\) sets for each instruction \(s\) for each iteration one would arrive at the two tables seen in Table \ref{tab:countll-rev-steps} and \ref{tab:countll-steps}.
The number of iterations is nearly halfed when reverseing the list of instructions prior to dataflow as can be seen on \ref{tab:countll-rev-steps} when compared to \ref{tab:countll-steps}.

@@

\begin{figure}[H]
     \centering
   \begin{minipage}[b]{0.50\linewidth}
     \centering
     \includesvg[height=10.9em]{countdown.briggs2}
     \caption{\texttt{build -{}- -t dot -a briggs -n 2}}\label{fig:countdown.briggs.dot2}
   \end{minipage}
   \begin{minipage}[b]{0.33\linewidth}
     \centering
     \includesvg[height=8.em]{countdown.briggs}
     \caption{\texttt{-a briggs -n 3}}\label{fig:countdown.briggs.dot}
   \end{minipage}
\end{figure}

@@

The primary text is \textit{Linear Scan Register Allocation} \cite{linear}, the application of which also relies on prior liveness analysis. Given liveranges, assignments are easily derived from only one pass through IR.
Similar to graph coloring, given \(k\) available GPRs, it assigns  as many variables as possible, and spilling the rest.

It does this by first building the intervals using the \(\mathit{in}[s]\) and \(\mathit{out}[s]\) resulting from prior dataflow analysis. This can be done in a single pass with a little creative array indexing as can be seen in Figure \ref{fig:intervals-ocaml}.

\begin{figure}[H]
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let intervals insns (in_, out) =
  let insn (starts, t, active) (i, n) =
    let init s (added, t, active) = (* .. *) in
    let end_ s (added, t, active) = (* .. *) in
    let a, t, c =
      init in_.(i) (S.SS.empty, t, active) |> init out.(i) |> end_ out.(i)
    in
    ((i, n, a) :: starts, t, c)
  in
  let starts, _, active =
    List.fold_left insn
      ([], S.ST.empty, Array.init (List.length insns) (Fun.const S.SS.empty))
      insns
  in
  List.rev starts |> List.map (fun (i, n, a) -> (i, n, a, active.(i)))
     \end{minted}
     \caption{Implementation of \texttt{intervals} in \texttt{linear.ml}}\label{fig:intervals-ocaml}
\end{figure}

\noindent The array indexing is used in the inner functions of \texttt{init} and \texttt{end\_} to unset the last point at which an interval goes out of scope.

\begin{figure}[H]
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let init s (added, t, active) =
  ( S.SS.filter (fun s -> not (S.ST.mem s t)) s |> S.SS.union added,
    S.SS.fold
      (fun s t ->
        S.ST.update s
          (function
            | Some j ->
                active.(i) <- S.SS.add s active.(i);
                active.(j) <- S.SS.remove s active.(j);
                Some i
            | None ->
                active.(i) <- S.SS.add s active.(i);
                Some i)
          t)
      s t,
    active )
     \end{minted}
     \caption{Implementation of \texttt{init} in \texttt{linear.ml}}\label{fig:init-ocaml}
\end{figure}

\begin{figure}[H]
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let end_ s (added, t, active) =
  ( S.SS.filter (fun s -> not (S.ST.mem s t)) s |> S.SS.union added,
    S.SS.fold
      (fun s t ->
        S.ST.update s
          (function
            | Some j ->
                active.(i) <- S.SS.add s active.(i);
                active.(j) <- S.SS.remove s active.(j);
                Some i
            | None -> Some i)
          t)
      s t,
    active )
     \end{minted}
     \caption{Implementation of \texttt{end\_} in \texttt{linear.ml}}\label{fig:end_-ocaml}
\end{figure}

\begin{figure}[H]
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let linearscan k insns =
  let avail = Regs.of_list (List.init k reg_of_int) in
  let insn (avail, active, assign) (_, _n, s, _e) =
    (*Printf.printf "%s\n" (Cfg.string_of_insn n);*)
    let start s (avail, active, assign) =
      match Regs.choose_opt avail with
      | Some reg -> (Regs.remove reg avail, S.ST.add s reg active, assign)
      | None ->
          (* FIXME: order by remaining or total length *)
          let z, reg = S.ST.choose active in
          ( avail,
            S.ST.remove z active |> S.ST.add s reg,
            S.ST.add z
              (Ind3 (Lit (Int64.of_int ((S.ST.cardinal assign * -8) - 8)), Rbp))
              assign )
    in
    S.SS.fold start s (avail, active, assign)
  in
  let _, active, assign =
    List.fold_left insn (avail, S.ST.empty, S.ST.empty) insns
  in
  S.ST.fold (fun k v acc -> S.ST.add k (Reg v) acc) active assign
     \end{minted}
     \caption{Implementation of \texttt{linearscan} in \texttt{linear.ml}}\label{fig:linearscan-ocaml}
\end{figure}
